Compositional Quantitative Resource Analysis of C Programs

### 2014-06-24 paper/main.tex
Q:Call
some changes to the description of the logic

Due to \code{break} and \code{return} statements of Clight, there are
different possible ways to exit a block of code.  We also have to keep
track of the resource specifications of functions.  To account for
this in the logic, our quantitative Hoare triples have the form
$$
\Delta; B; R \vdash
\htriple
  { Q }
  { S }
  { Q' } \; .
$$
The triple $\htriple{ Q }{ S }{ Q' }$ consists of a statement $S$ and
two assertions $Q,Q' : \Assn$.  It corresponds to triples in classic
Hoare logic and the intuitive meaning is as follows.  If $S$ is
executed with starting state $\state$, the empty continuation
$\Kstop$, and at least $P(\state)$ resources available then the
evaluation does not run out of resources and there are at least
$Q(\state')$ resources left if the evaluation terminates in state
$\state'$.  The assertion $B : \Assn$ provides the
postcondition for the case in which the code block $S$ is exited by a
\code{break} statement.  So if the execution is terminated in state
$\state'$ with a \code{break} then $B(\state')$ resources are
available.  Similarly, $R : \Z \to \Assn$ is the postcondition for the
case in which the code block $S$ is exited by a $\code{return}~x$
statement.  The integer argument of $R$ is the return value.
Finally, the function context of judgements that we write $\Delta$ is
a mapping from function names to specification of the form
$$
  \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v).
$$
The assertion $P_f \, z \, \vec v$ is the precondition of the
function $f$ and the assertion $Q_f \, z \, v$ is its postcondition.
They are both parameterized by an arbitary logical variable $z$
(which can be a tuple) that helps relating them.  The precondition
is allowed to depend on $\vec v$, the value of all the arguments at
the function invocation.  Similarly, the postcondition depends on
the return value $v$ of the function.

% The continuation-based semantics of a Clight requires that we
% distinguish pre- and postconditions in the   This is approach is
% standard in Hoare logics and followed for instance in Appel's
% separation logic for Clight~\cite{AppelLogic}.  Our postconditions
% $$
% Q = (Q^s, Q^b, Q^r) : \Assn \times \Assn \times (\Val \to \Assn)
% $$
% provide one assertion $Q^s$ for the case in which the block is exited by
% fall through, one assertion $Q^b$ if the block is exited by a \code{break}, and a
% function $Q^r$ from values to assertions in case the block is exited by a
% \code{return}.  The function argument in the last case represents the return
% value and the intended meaning is that the resulting assertion is
% guaranteed to hold for every return value.

% Since we have to deal with recursive functions, we also need a
% function context
% $$\Gamma {:} \funs {\to} ((\Val \times \Mem) {\to} \N \cup \{\infty\}) {\times} ((\Val \times \Mem) {\to} \N {\cup} \{\infty\})$$ 
% that maps function names to their specifications, that is, pre- and
% postconditions.  The precondition depends on the value that is passed
% to the function by the caller and the memory.  The postcondition
% depends on the return value and the memory.  For simplicity, we assume
% that a function has only one argument in this article.  In the Coq
% implementation, an arbitrary number of function arguments is allowed.

% In summary, a quantitative Hoare triple has the form
% $$
% \Gamma \vdash \htriple{P}{S}{Q}
% $$
% where $\Gamma$ is a function context, $P : \Assn$ is a precondition,
% $Q : \Assn \times \Assn \times (\Val \to \Assn)$ is a postcondition,
% and $S$ is a statement.

% Intuitively, an assertion can be seen as a \emph{potential function}
% that maps a program state to a non-negative potential.  The potential
% of the precondition $P$ must be sufficient to cover the cost of the
% execution of the statement $S$ and the potential $Q$ after the execution
% of $S$ (as in amortized resource analysis~\cite{HoffmannAH12}).

@@

\begin{figure*}[t]
\begin{mathpar}
%
\Rule{Q:Skip}
{}
{ \Gamma; B; R; Q \vdash \code{skip} \dashv Q; \Gamma }
%
\and \Rule{Q:Break}
{}
{ \Gamma; B; R; B + M_b \vdash \code{break} \dashv Q'; \Gamma' }
%
\and \Rule{Q:Return}
{}
{ \Gamma; B; R; R \, x + M_r \vdash \code{return}~x \dashv Q'; \Gamma' }
%
\and \Rule{Q:Assert}
{}
{ \Gamma; B; R; Q + M_a \vdash \code{assert}~e \dashv Q; \Gamma, e }
%
\and \Rule{Q:Loop}
{ \Gamma; Q'; R; Q \vdash S \dashv M_l + Q; \Gamma }
{ \Gamma; B; R; Q \vdash \code{loop}~S \dashv Q'; \code{inv}_S~\Gamma }
%
\and \Rule{Q:Set}
{ q'_{xy}, q'_{yx} \in \mathbb Q_0^+
\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
}
{ \Gamma[x/y]; B; R; Q + M_s \vdash x \gets y \dashv Q'; \Gamma }
%
\and \Rule{Q:IncrPos}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\}
\\\\ \textstyle q'_{0y} = q_{0y}
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{vx}
}
{ \textstyle
  \Gamma[x/x + y]; B; R; Q + M_s \vdash x \gets x + y \dashv Q'; \Gamma
}
%
\and \Rule{Q:IncrNeg}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
\\\\ \textstyle q'_{y0} = q_{y0}
      + \sum_{u \in \mathcal U} q_{ux}
      - \sum_{v \not\in \mathcal U} q_{xv}
}
{ \textstyle
  \Gamma[x/x - y]; B; R; Q + M_s \vdash x \gets x - y \dashv Q'; \Gamma
}
%
\and \Rule{Q:Incr}
{ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
}
{ \textstyle
  \Gamma[x/x \pm y]; B; R; Q + M_s \vdash x \gets x \pm y \dashv Q'; \Gamma
}
%
\and \Rule{Q:If}
{ \Gamma, e; B; R; Q - M_c^1 \vdash S_1 \dashv Q'; \Gamma'
\\ \Gamma, \neg e; B; R; Q - M_c^2 \vdash S_2 \dashv Q'; \Gamma'
}
{ \Gamma; B; R; Q \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv Q'; \Gamma' }
%
\and\Rule{Q:Seq}
{  \Gamma; B; R; Q - M_s^1 \vdash S_1 \dashv Q'; \Gamma'
\\ \Gamma'; B; R; Q' - M_s^2 \vdash S_2 \dashv Q''; \Gamma''
}
{ \Gamma; B; R; Q \vdash S_1;S_2 \dashv Q''; \Gamma'' }
%
\and\Rule{Q:Call}
{ (\vec y, s, \Gamma_f, P, P', \Gamma_f') \in \Delta(f)
\\ \ell = \text{Locals}(Q)
\\ R = P[\vec y / \vec x]
\\ \forall i \in \dom{R} . \, q_i = r_i
\\ R' = P'[s/r]
\\ \forall i \in \dom{R'} .\, q'_i = r'_i
\\ \forall i \not\in \dom{R'} .\, q'_i = 0
\\ \forall i \in \ell . \, q_i = q_i'
% \\ P' = Q'_f \Uparrow_{[r = s]}
% \\ \forall i \in \ell \,\setminus \{r\} .\, q_i = q_i'
% \\ \forall i \not\in \ell \,\setminus \{r\} .(q_i = p_i \land q'_i = p'_i)
% \\ c \in \mathbb Q_0^+
}
{ \Gamma_f[\vec y / \vec x] \land \Gamma_\ell; B; R; Q + c
  \vdash r \gets f(\vec x) \dashv
  Q' + c; \Gamma_f'[s/r] \land \Gamma_\ell
}
%
\and \Rule{Q:Weak}
{ \Gamma_1 \models \Gamma_2
\\ Q_1 \succeq_\Gamma Q_2
\\ \Gamma_2; B; R; Q_2 \vdash S \dashv Q'_2; \Gamma'_2
\\ Q'_2 \succeq_\Gamma Q'_1
\\ \Gamma'_2 \models \Gamma'_1
}
{ \Gamma_1; B; R; Q_1 \vdash S \dashv Q'_1; \Gamma'_1 }
%
\and \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy},~\Gamma \models l_{xy} \le |\inter x y| \}
\and \mathcal U = \{ xy \mid \exists u_{xy},~\Gamma \models |\inter x y| \le u_{xy} \}
\\ \forall i \in \mathcal U.\, q'_i \ge q_i - p_i
\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U \cup \mathcal L.\, q'_i \ge q_i
}
{ Q' + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i  \succeq_\Gamma Q }
%
\end{mathpar}
\caption{Abstract proof rules of the quantitative analysis}
\label{fig:proof}
\end{figure*}

>>>>

Due to \code{break} and \code{return} statements of Clight, there are
different possible ways to exit a block of code.  We also have to keep
track of the resource specifications of functions.  To account for
this in the logic, our quantitative Hoare triples have the form
$$
\Delta; B; R \vdash
\htriple
  { Q }
  { S }
  { Q' } \; .
$$
The triple $\htriple{ Q }{ S }{ Q' }$ consists of a statement $S$ and
two assertions $Q,Q' : \Assn$.  It corresponds to triples in classic
Hoare logic and the intuitive meaning is as follows.  If $S$ is
executed with starting state $\state$, the empty continuation
$\Kstop$, and at least $P(\state)$ resources available then the
evaluation does not run out of resources and there are at least
$Q(\state')$ resources left if the evaluation terminates in state
$\state'$.  The assertion $B : \Assn$ provides the
postcondition for the case in which the code block $S$ is exited by a
\code{break} statement.  So if the execution is terminated in state
$\state'$ with a \code{break} then $B(\state')$ resources are
available.  Similarly, $R : \Z \to \Assn$ is the postcondition for the
case in which the code block $S$ is exited by a $\code{return}~x$
statement.  The integer argument of $R$ is the return value.
Finally, the function context of judgements that we write $\Delta$ is
a mapping from function names to specification of the form
$$
  \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v).
$$
The assertion $P_f \, z \, \vec v$ is the precondition of the function
$f$ and the assertion $Q_f \, z \, v$ is its postcondition.  They are
both parameterized by an arbitary logical variable $z$ (which can be a
tuple) that relates the function arguments with the return value.  The
precondition also depends on $\vec v$, the value of all the arguments
at the function invocation.  Similarly, the postcondition depends on
the return value $v$ of the function.  The use of logical variables to
express relations between different states of an execution is a
standard technique of Hoare logic.

@@

\begin{figure*}[t]
\begin{mathpar}
%
\Rule{Q:Skip}
{}
{ \Gamma; B; R; Q \vdash \code{skip} \dashv Q; \Gamma }
%
\and \Rule{Q:Break}
{}
{ \Gamma; B; R; B + M_b \vdash \code{break} \dashv Q'; \Gamma' }
%
\and \Rule{Q:Return}
{}
{ \Gamma; B; R; R \, x + M_r \vdash \code{return}~x \dashv Q'; \Gamma' }
%
\and \Rule{Q:Assert}
{}
{ \Gamma; B; R; Q + M_a \vdash \code{assert}~e \dashv Q; \Gamma, e }
%
\and \Rule{Q:Loop}
{ \Gamma; Q'; R; Q \vdash S \dashv M_l + Q; \Gamma }
{ \Gamma; B; R; Q \vdash \code{loop}~S \dashv Q'; \code{inv}_S~\Gamma }
%
\and \Rule{Q:Set}
{ q'_{xy}, q'_{yx} \in \mathbb Q_0^+
\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
}
{ \Gamma[x/y]; B; R; Q + M_s \vdash x \gets y \dashv Q'; \Gamma }
%
\and \Rule{Q:IncrPos}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\}
\\\\ \textstyle q'_{0y} = q_{0y}
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{vx}
}
{ \textstyle
  \Gamma[x/x + y]; B; R; Q + M_s \vdash x \gets x + y \dashv Q'; \Gamma
}
%
\and \Rule{Q:IncrNeg}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
\\\\ \textstyle q'_{y0} = q_{y0}
      + \sum_{u \in \mathcal U} q_{ux}
      - \sum_{v \not\in \mathcal U} q_{xv}
}
{ \textstyle
  \Gamma[x/x - y]; B; R; Q + M_s \vdash x \gets x - y \dashv Q'; \Gamma
}
%
\and \Rule{Q:Incr}
{ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
}
{ \textstyle
  \Gamma[x/x \pm y]; B; R; Q + M_s \vdash x \gets x \pm y \dashv Q'; \Gamma
}
%
\and \Rule{Q:If}
{ \Gamma, e; B; R; Q - M_c^1 \vdash S_1 \dashv Q'; \Gamma'
\\ \Gamma, \neg e; B; R; Q - M_c^2 \vdash S_2 \dashv Q'; \Gamma'
}
{ \Gamma; B; R; Q \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv Q'; \Gamma' }
%
\and\Rule{Q:Seq}
{  \Gamma; B; R; Q - M_s^1 \vdash S_1 \dashv Q'; \Gamma'
\\ \Gamma'; B; R; Q' - M_s^2 \vdash S_2 \dashv Q''; \Gamma''
}
{ \Gamma; B; R; Q \vdash S_1;S_2 \dashv Q''; \Gamma'' }
%
\and\Rule{Q:Call}
{ (\vec y, s, \Gamma_f, Q_f, Q_f', \Gamma_f') \in \Delta(f)
\\ \text{Loc} = \text{Locals}(Q)
\\ \forall i \neq j . \, x_i \neq x_j
\\ c \in \mathbb Q_0^+
\\\\ Q \ge P + S
\\ P' + S \ge Q'
\\ R = Q_f[\vec y / \vec x]
\\ R' = Q'_f[s/r]
\\ \forall i \in \dom{R} . \, p_i = r_i
\\ \forall i \in \dom{R'} .\, p'_i = r'_i 
\\ \forall i \not\in \dom{R'} .\, p'_i = 0
\\ \forall i \not\in \text{Loc}
  .\, s_i = 0
}
{ \Gamma_f[\vec y / \vec x] \land \Gamma_{\text{Loc}}; B; R; Q + c
  \vdash r \gets f(\vec x) \dashv
  Q' + c; \Gamma_f'[s/r] \land \Gamma_{\text{Loc}}
}
%
\and \Rule{Q:Weak}
{ \Gamma_1 \models \Gamma_2
\\ Q_1 \succeq_\Gamma Q_2
\\ \Gamma_2; B; R; Q_2 \vdash S \dashv Q'_2; \Gamma'_2
\\ Q'_2 \succeq_\Gamma Q'_1
\\ \Gamma'_2 \models \Gamma'_1
}
{ \Gamma_1; B; R; Q_1 \vdash S \dashv Q'_1; \Gamma'_1 }
%
\and \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy},~\Gamma \models l_{xy} \le |\inter x y| \}
\and \mathcal U = \{ xy \mid \exists u_{xy},~\Gamma \models |\inter x y| \le u_{xy} \}
\\ \forall i \in \mathcal U.\, q'_i \ge q_i - p_i
\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U \cup \mathcal L.\, q'_i \ge q_i
}
{ Q' + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i  \succeq_\Gamma Q }
%
\end{mathpar}
\caption{Abstract proof rules of the quantitative analysis}
\label{fig:proof}
\end{figure*}
